Using Heavy-Tailed Self-Regularization Theory to Characterize Correlations in and Perform Diagnostics on State-of-the-art Overparameterized Models

### 2021-02-18 submission/TOPML_Workshop/paper.tex
work on simpsons section, its pretty mature now

We find that norm based metrics correlate well with reported test accuracies for well-trained models, but that they often cannot distinguish well-trained versus poorly-trained models.
We also find that power law based metrics can do much better---quantitatively better at discriminating among series of well-trained models with a given architecture; and qualitatively better at discriminating well-trained versus poorly-trained models.
These methods can be used to identify when a pretrained neural network has problems that cannot be detected simply by examining training/test accuracies.
%
%Here is a summary of our main results.
%First, norm-based metrics do a reasonably good job at predicting quality trends in well-trained CV/NLP models.
%Second, norm-based metrics may give spurious results when applied to poorly-trained models (e.g., models trained without enough data, etc.).
%For example, they may exhibit what we call Scale Collapse for these models. 
%Third, PL-based metrics can do much better at predicting quality trends in pretrained CV/NLP models.  
%In particular, 
%a weighted PL exponent (weighted by the log of the spectral norm of the corresponding layer) is 
%quantitatively better at discriminating among a series of well-trained versus very-well-trained models
%within a given architecture series; and
%the (unweighted) average PL exponent is 
%qualitatively better at discriminating well-trained versus poorly-trained models.
%Fourth, PL-based metrics can also be used to characterize fine-scale model properties, including what we call layer-wise Correlation Flow, in well-trained and poorly-trained models; and they can be used to evaluate model enhancements (e.g., distillation, fine-tuning, etc.).
%
Our work provides a theoretically-principled empirical evaluation---by far the largest, most detailed, and most comprehensive to date---and it uses HT-SR theory.%
\footnote{Performing such a meta-analysis of previously-published work is common in certain areas, but it is quite rare in ML, where the emphasis is on developing better training protocols.}

%\citep{BFT17_TR} is Bartlett: ``Spectrally-normalized margin bounds for neural networks''
%\citep{LMBx18_TR} is Poggio: ``A surprising linear relationship predicts test performance in deep networks''
There is related work \citep{BFT17_TR,LMBx18_TR}, but that work provides generalization bounds, modifies the models, requires access to data, etc., and so is quite different.

@@

XXX.  IF WANT A ONE SIZE FITS ALL METRIC, THEN CAN EXPECT A SIMPSONS PARADOX.  OUR TECHNIQUES CAN DIAGNOST THIS.  AND IN FACT WE SEE THAT.

\michael{Charles, put in a paragraph and the figure for the Simpson's paradox here.}

\charles{We apply our theory of Heavy Tailed Self-Regularization (HTSR) to predict the generalization accuracies of the pretrained models provided in this contest, as a function of both model depth (L) and the solver hyperparameters (batch size, dropout, weight decay), and compare quality metrics that describe both the heavy-tailed shape and norm-based scale of the model layer weight matrices.  In doing so, we identify what amounts to a Simpson’s paradox, in which
shape and scale metrics show opposite behavior, and a combine metric resolves the paradox}

\begin{figure}[t!] 
    \centering
    \subfigure[Log Spectral Norm]{
       \includegraphics[width=3.5cm]{img/simpsons-task2-logspectralnorm.png}
    }
    \subfigure[Weighted Alpha $\hat{\alpha}$]{
       \includegraphics[width=3.5cm]{img/simpsons-task2-alpha_weighted.png}
    }
    \caption{Test accuracy as a function of 
             the Log Spectral Norm and 
             the (HT-SR) Weighted Alpha ($\hat{\alpha}$) metrics,
             as model hyperparameters and depth are varied. 
             Models are from the NeurIPS Predicting Generalization contest.
             A Simpson's paradox is clearly visible.
            }
    \label{fig:simpsons}
\end{figure}

\citep{JNBx19_fantastic_TR} is the terrible fantastic paper: ``Fantastic Generalization Measures and Where to Find Them''

\citep{JFYx20_contest_v10} is the contest: ``{NeurIPS} 2020 Competition: Predicting Generalization in Deep Learning (Version 1.0)''

\citep{BHO75} is Bickel: ``Sex Bias in Graduate Admissions: Data from Berkeley''

@@

\michael{I may cut this stuff.}
We have shown that weight matrix analysis derived from HT-SR Theory can be used to predict trends in the quality of state-of-the-art neural networks without access to training or testing data as well as identify Simpson's-like paradoxes in publicly-available NN models.
The HT-SR Theory underlying our analysis is quite different than theoretical approaches popular in ML, and indeed it is quite different than typical statistical mechanics approaches to NNs \citep{BKPx20}.
Recently, however, work motivated by HT-SR Theory has 
demonstrated that multiplicative noise in stochastic optimization leads to heavy tails~\citep{HodMah20A_TR,GSZ20_TR}; 
provided a random matrix analysis of random {F}ourier features, going beyond the {G}aussian kernel, to provide a precise phase transition and illustrate the corresponding double descent~\citep{ZCM20_TR}; 
provided exact expressions for double descent and implicit regularization via a notion of surrogate random design \citep{DLM19_Exact_TR}; 
demonstrated that good classifiers are abundant in the interpolating regime \citep{TKM20_TR};
shown how injecting noise into hidden states of recurrent NNs leads to implicit regularization~\citep{LEHM21_TR,Mah12}; and 
shown that analogous phase transitions hold for the {C}olumn {S}ubset {S}election and the {N}ystrom method~\citep{DKM20_TR}. 
We expect that the empirical results identified by tools from HT-SR Theory will provide guidance and constraints for future development of overparameterized machine learning models.

>>>>

We find that norm based metrics correlate well with reported test accuracies for well-trained models, but that they often cannot distinguish well-trained versus poorly-trained models.
We also find that PL based metrics can do much better---quantitatively better at discriminating among series of well-trained models with a given architecture; and qualitatively better at discriminating well-trained versus poorly-trained models.
These methods can be used to identify when a pretrained DNN has problems that cannot be detected simply by examining training/test accuracies.
%
%Here is a summary of our main results.
%First, norm-based metrics do a reasonably good job at predicting quality trends in well-trained CV/NLP models.
%Second, norm-based metrics may give spurious results when applied to poorly-trained models (e.g., models trained without enough data, etc.).
%For example, they may exhibit what we call Scale Collapse for these models. 
%Third, PL-based metrics can do much better at predicting quality trends in pretrained CV/NLP models.  
%In particular, 
%a weighted PL exponent (weighted by the log of the spectral norm of the corresponding layer) is 
%quantitatively better at discriminating among a series of well-trained versus very-well-trained models
%within a given architecture series; and
%the (unweighted) average PL exponent is 
%qualitatively better at discriminating well-trained versus poorly-trained models.
%Fourth, PL-based metrics can also be used to characterize fine-scale model properties, including what we call layer-wise Correlation Flow, in well-trained and poorly-trained models; and they can be used to evaluate model enhancements (e.g., distillation, fine-tuning, etc.).
See Table~\ref{table:results} for an overview summary of our results.

Our work provides a theoretically-principled empirical evaluation---by far the largest, most detailed, and most comprehensive to date---and it uses HT-SR theory.
%
%\citep{BFT17_TR} is Bartlett: ``Spectrally-normalized margin bounds for neural networks''
%\citep{LMBx18_TR} is Poggio: ``A surprising linear relationship predicts test performance in deep networks''
There is related work \citep{BFT17_TR,LMBx18_TR}, but that work provides generalization bounds, modifies the models, requires access to data, etc., and so is quite different.

@@

%\citep{JNBx19_fantastic_TR} is the terrible fantastic paper: ``Fantastic Generalization Measures and Where to Find Them''
%
%\citep{JFYx20_contest_v10} is the contest: ``{NeurIPS} 2020 Competition: Predicting Generalization in Deep Learning (Version 1.0)''
%
%\citep{BHO75} is Bickel: ``Sex Bias in Graduate Admissions: Data from Berkeley''

\begin{figure}[t!] 
    \centering
    \subfigure[Log Spectral Norm]{
       \includegraphics[width=3.5cm]{img/simpsons-task2-logspectralnorm.png}
    }
    \subfigure[Weighted Alpha $\hat{\alpha}$]{
       \includegraphics[width=3.5cm]{img/simpsons-task2-alpha_weighted.png}
    }
    \caption{Test accuracy as a function of 
             the Log Spectral Norm and 
             the (HT-SR) Weighted Alpha ($\hat{\alpha}$) metrics,
             as model hyperparameters and depth are varied. 
             Models are from the NeurIPS 2020 ``Predicting Generalization'' contest.
             A Simpson's paradox is clearly visible.
            }
    \label{fig:simpsons}
\end{figure}

Subsequent to the development of HT-SR theory, and our initial results on analyzing pretrained models at scale, there was at NeurIPS 2020 a ``Predicting Generalization'' workshop and contest \citep{JNBx19_fantastic_TR, JFYx20_contest_v10}.
This contest made publicly-available pretrained models.
This corpus of models was much more narrow (in terms of domain) but more extensive (in terms of depth and hyperparameter variation) than the previous SOTA corpus we considered~\citep{osmr}.

In particular, we apply HT-SR theory to predict the generalization accuracies of the pretrained models provided in this contest, as a function of both model depth ($L$) and the solver hyperparameters (batch size, dropout, weight decay), and we compare quality metrics that describe the norm-based scale and the HT/PL shape of the model layer weight matrices.  
See Figure~\ref{fig:simpsons} for a summary.
For norm-based metrics, we identify what amounts to a Simpson’s paradox \citep{BHO75}, in which scale and shape metrics show opposite behavior, depending on whether the data are partitioned by depth or solver hyperparameters values.
Since $\hat{\alpha}$ combines both scale and shape information, it resolves the paradox.

%\charles{We apply our theory of Heavy Tailed Self-Regularization (HTSR) to predict the generalization accuracies of the pretrained models provided in this contest, as a function of both model depth (L) and the solver hyperparameters (batch size, dropout, weight decay), and compare quality metrics that describe both the heavy-tailed shape and norm-based scale of the model layer weight matrices.  In doing so, we identify what amounts to a Simpson’s paradox, in which shape and scale metrics show opposite behavior, and a combine metric resolves the paradox}

@@

The HT-SR theory underlying our analysis is quite different than theoretical approaches popular in machine learning, and it is quite different than typical statistical mechanics approaches to DNNs \citep{BKPx20}.
%
%Recently, however, work motivated by HT-SR theory has 
%demonstrated that multiplicative noise in stochastic optimization leads to heavy tails~\citep{HodMah20A_TR,GSZ20_TR}; 
%provided a random matrix analysis of random {F}ourier features, going beyond the {G}aussian kernel, to provide a precise phase transition and illustrate the corresponding double descent~\citep{ZCM20_TR}; 
%provided exact expressions for double descent and implicit regularization via a notion of surrogate random design \citep{DLM19_Exact_TR}; 
%demonstrated that good classifiers are abundant in the interpolating regime \citep{TKM20_TR};
%shown how injecting noise into hidden states of recurrent NNs leads to implicit regularization~\citep{LEHM21_TR,Mah12}; and 
%shown that analogous phase transitions hold for the {C}olumn {S}ubset {S}election and the {N}ystrom method~\citep{DKM20_TR}. 
%We expect that the empirical results identified by tools from HT-SR theory will provide guidance and constraints for future development of overparameterized machine learning models.
%
However, recent work, work motivated by HT-SR theory, has begun to appear
\citep{HodMah20A_TR,GSZ20_TR, ZCM20_TR, DLM19_Exact_TR, TKM20_TR, LEHM21_TR,Mah12, DKM20_TR}. 
We expect that the empirical results and theoretical insights HT-SR theory provides will provide guidance and constraints for future development of overparameterized machine learning models.
